Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-06-01 manuscript.tex
zotero update

In this article, we focus on the role of motivation and need fulfillment to understand when and why exactly an interaction is perceived as positive. We propose need fulfillment in particular because needs are a fundamental aspect of the human experience that governs a significant number of emotional, cognitive, and behavioral facets \citep[][]{kreienkamp2022d, kruglanski2002}. Importantly, need fulfillment has particularly been highlighted in explaining the success of (close) relationships, psychosocial functioning, as well as reducing conflict between groups --- all of which are essential to positive intergroup interactions.

@@

% Motivation and core needs 
A broader theoretical implication relates to the role of situational motivation in intergroup contacts. Our results offer a first promising test of a psychological mechanism of need fulfillment in intergroup contact. While our results are tentative given their novelty within the field, they were highly consistent across studies and may offer new theoretical avenues. Experiences of need fulfillment are a facet of the human experience that has thus far been underemphasized in the intergroup contact literature. This stands in stark contrast to the many cognitive \citep[e.g.,][]{Pettigrew1998, Brown2005} and emotional aspects investigated within the field \citep[e.g.,][]{Stephan2008}. Future research may, therefore, be able to integrate broader theoretical frameworks of intergroup contact (e.g., motivations guiding cognition and affect, which in turn drive behavior. cf., theory of reasoned goal pursuit; \citealp{Ajzen2019}. Also see \citealp{kreienkamp2022d}).

@@

\paragraph{Embeddedness of code needs}
We, finally, use the qualitative data from the participants' self-identified core needs to contextualize our results. We leverage machine learning to extract a topic model of the free-text entries across the three studies. We describe the extracted topics and themes and compare them to the need contents usually found and measured within the psychological literature. Full methodological details and visualizations are available in Supplemental Material B.

>>>>

In this article, we focus on the role of motivation and need fulfillment to understand when and why exactly an interaction is perceived as positive. We propose need fulfillment in particular because needs are a fundamental aspect of the human experience that governs a significant number of emotional, cognitive, and behavioral facets \citep[][]{Kreienkamp2022d, kruglanski2002}. Importantly, need fulfillment has particularly been highlighted in explaining the success of (close) relationships, psychosocial functioning, as well as reducing conflict between groups --- all of which are essential to positive intergroup interactions.

@@

% Motivation and core needs 
A broader theoretical implication relates to the role of situational motivation in intergroup contacts. Our results offer a first promising test of a psychological mechanism of need fulfillment in intergroup contact. While our results are tentative given their novelty within the field, they were highly consistent across studies and may offer new theoretical avenues. Experiences of need fulfillment are a facet of the human experience that has thus far been underemphasized in the intergroup contact literature. This stands in stark contrast to the many cognitive \citep[e.g.,][]{Pettigrew1998, Brown2005} and emotional aspects investigated within the field \citep[e.g.,][]{Stephan2008}. Future research may, therefore, be able to integrate broader theoretical frameworks of intergroup contact (e.g., motivations guiding cognition and affect, which in turn drive behavior. cf., theory of reasoned goal pursuit; \citealp{Ajzen2019}. Also see \citealp{Kreienkamp2022d}).

@@

\paragraph{Embeddedness of code needs}
We, finally, use the qualitative data from the participants' self-identified core needs to contextualize our results. We leverage machine learning to extract a topic model of the free-text entries across the three studies. We describe the extracted topics and themes and compare them to the need contents usually found and measured within the psychological literature. Full methodological details and visualizations are available in Supplemental Material C.
